Enhancing diagnostic of stochastic mortality models leveraging contrast trees. An application on Italian data

### 2022-11-11 Manuscript Contrast Trees.tex
andrea

%The life expectancy rise that has occurred in the last decades constitutes a great conquest but contemporarily creates substantial challenges for life insurers and social security schemes providing benefits related to illness, or survival of people. Important policy decisions are based on the human longevity reliable estimates, and public pension policies are a prime example. Thereby, understanding longevity dynamics over age and time is crucial, as they are one of the driving forces of population change. 
Since 1980, innovative approaches and developments in mortality modeling have been constantly proposed. Mortality analysis has received a considerable contribution from statistical science, building solid foundations for the evolution of mortality methods. Estimating longevity is not straightforward; accuracy depends on the particular situation or trends, but it is not clear when a method will perform best. Indeed, new mortality forecasting techniques will appear in the literature but may take years before they can be fully evaluated. As proposed by \cite{Booth}, the accuracy of population estimates should be regularly tested establishing the improvement evidence. Researchers appear to be more focused on technical progress in methods rather than on the accuracy of the estimation they provide, in particular minimizing the bias.

Several approaches have been used to model mortality surface, determining how death rates change over time. Until the 1980s, mortality models were relatively simple and involved a fair degree of subjective judgment (for a review see \cite{Pollard}). The growing availability of reliable data, in lockstep with the improvement of statistical-mathematical methods, has allowed the creation of ever-finer mortality forecasting models. According to \cite{Booth}, literature would suggest three approaches to demographic modeling. The first one (explanation) makes use of structural or epidemiological models from certain causes of death. A classic example is the dependence of lung cancer on tobacco smoking. The second one (expectation) is based on subjective expert opinion, involving varying degrees of formality. Finally, the third and most commonly used approach is extrapolative, using the regularity typically found in both age patterns and trends over time.
This approach includes the more complex stochastic models such as the Lee-Carter \cite{LC92} and, more in general, the Generalized Age Period Cohort (GAPC) model. Despite the Lee-Carter model being widely recognized as the cornerstone of mortality modeling and forecasting, over the last decade, scholars have suggested additional approaches that have also gained interest in the academic world \cite{BDV2002,RH,CBD06,CBD09}.

@@

The significant reduction in the forecasting error reached by the application of machine learning techniques became particularly useful for both researchers and practitioners. The main contributions are from \cite{Deprez}, \cite{LevPiz} and \cite{LevNi}. The common idea behind all these works is to improve the fitting accuracy of canonical models using machine learning algorithms. In other words, to correct the mortality surface produced by standard stochastic mortality models. All of the proposed methods, calibrate a machine learning estimator used to adjust (and improve) the mortality rates estimated by the original mortality model. Those authors show that mortality modeling can benefit from machine learning that better captures patterns that traditional models do not identify. 
%Specifically, \cite{LevNi} proposed to extrapolate the machine learning estimator on the whole mortality surface using the two-dimensional P-splines, obtaining more accurate mortality projection with respect to the Lee-Carter model.

The need for new models that can understand mortality improvements more accurately than canonical tools is evident. 
This paper contributes to the literature on mortality evaluation by introducing an innovative approach based on machine learning techniques that demographers have not yet explored, contributing to the undervalued field of models assessment. 
Specifically, we provide an application of the method proposed by \cite{Friedman2020}, namely Contrast Trees, for evaluating the accuracy of the mortality estimates given by different models that are not treatable with standard validation methods (e.g. those based on the likelihood function). 
The key significance of the Contrast Trees in mortality evaluation is the supply of a unified approach allowing for assessing and comparing the accuracy of both standard mortality models and machine learning algorithms involved in mortality estimates. Moreover, in addition to evaluating the accuracy of the models, the Contrast Trees enables improving the performance of the models through a boosting procedure that reduces the inaccuracies. We use this methodology, namely Contrast Boosting, to improve the mortality rates estimates.

@@

%-------------------------------------------------------------
\section*{Discussion}
%-------------------------------------------------------------
%\textit{The Discussion should be succinct and must not contain subheadings.}\\

CTs consist of a general method based on machine learning that can be applied to any model, expressed as a regression model, to evaluate the goodness of fit and identify the worst-performing regions in the input space. While, other well-known goodness of fit evaluation criteria, such as Bayesian Information Criterion, Akaike Information Criterion, Schwartz Information Criterion, require the likelihood function, which is not available for machine learning models. Therefore, it provides a unified approach for assessing and comparing the accuracy of both traditional models and machine learning algorithms.

>>>>

%The life expectancy rise that has occurred in the last decades constitutes a great conquest but contemporarily creates substantial challenges for life insurers and social security schemes providing benefits related to illness, or survival of people. Important policy decisions are based on the human longevity reliable estimates, and public pension policies are a prime example. Thereby, understanding longevity dynamics over age and time is crucial, as they are one of the driving forces of population change. 
Since 1980, innovative approaches and developments in mortality modeling have been constantly proposed. Mortality analysis has received a considerable contribution from statistical science, building solid foundations for the evolution of mortality methods. Estimating longevity is not straightforward; accuracy depends on the particular situation or trends, but it is not clear when a method will perform best. Indeed, new mortality models will appear in the literature but may take years before they can be fully evaluated. As proposed by \cite{Booth}, the accuracy of population estimates should be regularly tested establishing the improvement evidence. Researchers appear to be more focused on technical progress in methods rather than on the accuracy of the estimation they provide, in particular minimizing the bias.

Several approaches have been used to model mortality surface, determining how death rates change over time. Until the 1980s, mortality models were relatively simple and involved a fair degree of subjective judgment (for a review see \cite{Pollard}). The growing availability of reliable data, in lockstep with the improvement of statistical-mathematical methods, has allowed the creation of ever-finer mortality models. According to \cite{Booth}, literature would suggest three approaches to demographic modeling. The first one (explanation) makes use of structural or epidemiological models from certain causes of death. A classic example is the dependence of lung cancer on tobacco smoking. The second one (expectation) is based on subjective expert opinion, involving varying degrees of formality. Finally, the third and most commonly used approach is extrapolative, using the regularity typically found in both age patterns and trends over time.
This approach includes the more complex stochastic models such as the Lee-Carter \cite{LC92} and, more in general, the Generalized Age Period Cohort (GAPC) model. Despite the Lee-Carter model being widely recognized as the cornerstone of mortality modeling and forecasting, over the last decade, scholars have suggested additional approaches that have also gained interest in the academic world \cite{BDV2002,RH,CBD06,CBD09}.

@@

The significant reduction in the estimation error reached by the application of machine learning techniques became particularly useful for both researchers and practitioners. The main contributions are from \cite{Deprez}, \cite{LevPiz} and \cite{LevNi}. The common idea behind all these works is to improve the fitting accuracy of canonical models using machine learning algorithms. In other words, to correct the mortality surface produced by standard stochastic mortality models. All of the proposed methods, calibrate a machine learning estimator used to adjust (and improve) the mortality rates estimated by the original mortality model. Those authors show that mortality modeling can benefit from machine learning that better captures patterns that traditional models do not identify. 
%Specifically, \cite{LevNi} proposed to extrapolate the machine learning estimator on the whole mortality surface using the two-dimensional P-splines, obtaining more accurate mortality projection with respect to the Lee-Carter model.

The need for new models that can understand mortality improvements more accurately than canonical tools is evident. 
This paper contributes to the literature on mortality evaluation by introducing an innovative approach based on machine learning techniques that demographers have not yet explored, contributing to the undervalued field of models assessment. 
Specifically, we provide an application of the method proposed by \cite{Friedman2020}, namely Contrast Trees, for evaluating the accuracy of the mortality estimates given by different models that are not treatable with standard validation methods (e.g. those based on the likelihood function). 
The key significance of the Contrast Trees in mortality evaluation is the supply of a unified approach allowing for assessing and comparing the accuracy of both standard mortality models and machine learning algorithms involved in mortality estimates. Moreover, in addition to evaluating the accuracy of the models, the Contrast Trees enables improving the performance of the models through a boosting procedure that reduces the inaccuracies. We use this methodology, namely Contrast Boosting, to improve the mortality rates estimates. \textcolor{blue}{Indeed, according to the demographic literature, the reliable estimation of mortality data may refer not only to the extrapolation but also to an accurate fitting of the historical mortality surface.  
For instance, in longevity analysis is common to deal with subpopulations i.e. regions or provinces, characterized by a high level of stochasticity often due to a small number of count data at single ages. This is the case in which specific ages or years are not covered with data information, making the mortality estimation challenging. Our approach is crucial to evaluate the mortality matrix estimation provided by a mortality model, to ensure the effectiveness of estimation comparing  different methods leveraging the proposed approach}

@@

%-------------------------------------------------------------
\section*{Discussion}
%-------------------------------------------------------------
%\textit{The Discussion should be succinct and must not contain subheadings.}\\
\textcolor{blue}{Evaluate and thus eventually improve the mortality models fit is crucial for both demographers and actuaries. Indeed, in particular situations, common in actuarial practice, data quality can turn the mortality estimate difficult. A prime example is the case of small subpopulations where the common method such as LC may not guarantee reliable estimation. In this sense, our proposal fills the gap between mortality modeling and model diagnostics, in particular for nontraditional modeling as a machine learning framework }
CTs consist of a general method based on machine learning that can be applied to any model, expressed as a regression model, to evaluate the goodness of fit and identify the worst-performing regions in the input space. While, other well-known goodness of fit evaluation criteria, such as Bayesian Information Criterion, Akaike Information Criterion, Schwartz Information Criterion, require the likelihood function, which is not available for machine learning models. Therefore, it provides a unified approach for assessing and comparing the accuracy of both traditional models and machine learning algorithms.
